---
title: "Weather Damage Report"
author: "Victor Wildner"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, cache = TRUE)
```

> The basic goal of this project is to explore the U.S. National Oceanic and Atmospheric Administration’s (NOAA) storm database, seeking to answer some questions about severe weather events. The main focus of this study is to enable full-research reproducibility, using literate programming for every step.


## Data Processing

### Setup

#### Loading libraries

```{r libs, echo=FALSE, results=FALSE}
library(tidyverse)
library(cowplot)
```

#### Setting seed for reproducibility

```{r seed}
set.seed(144)
```

#### Showing current environment

```{r sessioninfo}
sessionInfo()
```


#### Downloading the dataset

```{r download, cache = TRUE}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2", "stormdata.csv.bz2")
```

#### Read dataset into variable removing blank spaces and transform into tidyverse compatible

```{r load_data, cache = TRUE}
storm_data <- as_tibble(read.csv(file = "stormdata.csv.bz2", na.strings = ""))
```

#### Get a glimpse of the data
```{r glimpse}
glimpse(storm_data)
```

#### Count NAs

```{r countnas}
sum_na <- storm_data %>%
    summarize_all(funs(sum(is.na(.))))
glimpse(sum_na)
```

#### Select only needed data from the processed dataset

```{r filterdata}
filter_data <- storm_data %>% 
    select(STATE, EVTYPE, FATALITIES, INJURIES, PROPDMG, PROPDMGEXP, CROPDMG, CROPDMGEXP)

glimpse(filter_data)
```

#### Impute 0 to NAs

```{r tidyone}
tidy_storm <- filter_data %>%
    replace(is.na(.), "0"
    ) %>% 
    mutate(PROPDMGEXP = as.character(PROPDMGEXP),
           CROPDMGEXP = as.character(CROPDMGEXP))
glimpse(tidy_storm)
```

#### Find unique values in prop and crop

```{r findunique}
unique_prop <- unique(tidy_storm$PROPDMGEXP)
unique_crop <- unique(tidy_storm$CROPDMGEXP)

unique_prop
unique_crop
```

#### Create dictionary for the unique values

```{r dictionary}
rep_prop <- c(1e3, 1e6, 0, 1e9, 1e6, 0, 1e5, 1e6, 0, 1e4, 1e2, 1e3, 1e2, 1e7, 1e2, 0, 1e1, 1e8)
rep_crop <- c(0, 1e6, 1e3, 1e6, 1e9, 0, 1e3, 1e2)

prop_dictionary <- cbind(unique_prop, rep_prop)
crop_dictionary <- cbind(unique_crop, rep_crop)
prop_dictionary
crop_dictionary
```

#### Remap values back into the tidy data set using plyr


```{r tidy}
tidy_storm$PROPDMGEXP <- plyr::mapvalues(tidy_storm$PROPDMGEXP, from = unique_prop, to = rep_prop)
tidy_storm$CROPDMGEXP <- plyr::mapvalues(tidy_storm$CROPDMGEXP, from = unique_crop, to = rep_crop)

tidy_storm <- tidy_storm %>% 
    mutate(PROPDMGEXP = as.double(PROPDMGEXP),
           CROPDMGEXP = as.double(CROPDMGEXP))

glimpse(tidy_storm)
```

## Questions

### 1. Across the United States, which types of events (as indicated in the EVTYPE variable are most harmful with respect to population health?

#### Rank fatalities and injuries according to EVTYPE

```{r nowquestion}
sum_events_global <- tidy_storm %>%
    select(-STATE) %>% 
    group_by(EVTYPE) %>%
    summarize_all(
        list(sum)
    )

sum_fatalities_global <- sum_events_global %>% 
    select(EVTYPE, FATALITIES) %>% 
    arrange(desc(FATALITIES))

sum_fatalities_top10 <- slice(sum_fatalities_global, 1:10)

sum_injuries_global <- sum_events_global %>% 
    select(EVTYPE, INJURIES) %>% 
    arrange(desc(INJURIES))

sum_injuries_top10 <- slice(sum_injuries_global, 1:10)

```

#### Plot results

```{r plot1, warning=FALSE}
plot1 <- ggplot(sum_fatalities_top10, aes(reorder(EVTYPE, -FATALITIES), FATALITIES, group = EVTYPE, alpha = EVTYPE)) +
    geom_bar(stat = "identity") +
    theme_minimal() +
    theme(
        axis.text.x = element_text(angle = 90),
        legend.position = "none"
          ) +
    labs(
        title = "Major Causes of Fatality",
        x = "",
        y = "Fatalities"
    )

plot2 <- ggplot(sum_injuries_top10, aes(reorder(EVTYPE, -INJURIES), INJURIES, group = EVTYPE, alpha = EVTYPE, fill = "red")) +
    geom_bar(stat = "identity") +
    theme_minimal() +
    theme(
        axis.text.x = element_text(angle = 90),
        legend.position = "none"
          ) +
    labs(
        title = "Major Causes of Injuries",
        x = "",
        y = "Injuries"
    )
plot_grid(plot1, plot2)
```


### 2. Across the United States, which types of events have the greatest economic consequences?

### Prepare vars for measured and expected damage

```{r measurednexpected}

# Measured damage

propdmg_global <- sum_events_global %>% 
    select(EVTYPE, PROPDMG) %>% 
    arrange(desc(PROPDMG))

propdmg_top10 <- slice(propdmg_global, 1:10)

cropdmg_global <- sum_events_global %>% 
    select(EVTYPE, CROPDMG) %>% 
    arrange(desc(CROPDMG))

cropdmg_top10 <- slice(cropdmg_global, 1:10)

# Expected damage
exp_propdmg_global <- sum_events_global %>% 
    select(EVTYPE, PROPDMGEXP) %>% 
    arrange(desc(PROPDMGEXP))

exp_propdmg_top10 <- slice(exp_propdmg_global, 1:10)

exp_cropdmg_global <- sum_events_global %>% 
    select(EVTYPE, CROPDMGEXP) %>% 
    arrange(desc(CROPDMGEXP))

exp_cropdmg_top10 <- slice(exp_cropdmg_global, 1:10)

```
#### Plot the results 

```{r plot2, warning=FALSE}
plot3 <- ggplot(propdmg_top10, aes(reorder(EVTYPE, -PROPDMG), PROPDMG, group = EVTYPE, alpha = EVTYPE)) +
    geom_bar(stat = "identity", fill = "rosybrown") +
    theme_minimal() +
    theme(
        axis.text.x = element_text(angle = 90),
        legend.position = "none"
          ) +
    labs(
        title = "Major Causes of Prop Damage",
        x = "",
        y = "Gross cost ($)"
    )

plot4 <- ggplot(cropdmg_top10, aes(reorder(EVTYPE, -CROPDMG), CROPDMG, group = EVTYPE, alpha = EVTYPE)) +
    geom_bar(stat = "identity", fill = "navyblue") +
    theme_minimal() +
    theme(
        axis.text.x = element_text(angle = 90),
        legend.position = "none"
          ) +
    labs(
        title = "Major Causes of Crop Damage",
        x = "",
        y = "Gross cost ($)"
    )
plot_grid(plot3, plot4)
```
```{r plot3, warning=FALSE}
plot5 <- ggplot(exp_propdmg_top10, aes(reorder(EVTYPE, -PROPDMGEXP), PROPDMGEXP, group = EVTYPE, alpha = EVTYPE)) +
    geom_bar(stat = "identity", fill = "firebrick") +
    theme_minimal() +
    theme(
        axis.text.x = element_text(angle = 90),
        legend.position = "none"
          ) +
    labs(
        title = "Expected Causes of Prop Damage",
        x = "",
        y = "Gross cost ($)"
    )

plot6 <- ggplot(exp_cropdmg_top10, aes(reorder(EVTYPE, -CROPDMGEXP), CROPDMGEXP, group = EVTYPE, alpha = EVTYPE)) +
    geom_bar(stat = "identity", fill = "slateblue3") +
    theme_minimal() +
    theme(
        axis.text.x = element_text(angle = 90),
        legend.position = "none"
          ) +
    labs(
        title = "Expected Causes of Crop Damage",
        x = "",
        y = "Gross cost ($)"
    )
plot_grid(plot5, plot6, align = "hv")
```


## Results

The extreme events that took most lives and caused most injuries in the United States are tornado, heat, and flash flood with tornado being about 2/3 of the total reported cases.

The events that had the most economic damage in the US are hurricanes, tornadoes, flood and drought.
